Quantitative Evaluation of Touch-Screen Group Musical Performance with Transition Matrix Measures

### 2015-09-15 flux-entropy-paper/flux-entropy-paper.tex
boom
first complete draft

\title{Quantitative Evaluation of Touch-Screen Performance with
  Transition Matrix Measures}

@@

\begin{abstract}
  HCI systems for supporting collaborative creativity are often
  evaluated with user ratings and interviews. However, such data can
  be impractical to collect in-the-wild, for example during musical
  performances. In this paper we show how a transition matrix based
  post-hoc analysis of logged interaction from a long program of
  experiments in group musical creativity can shed light on the nature
  of the group interaction in-the-wild. Using a corpus of more than
  140 performances, rehearsals, and demonstrations over more than two
  years, we find that two matrix measures, flux and entropy, can be
  used as a proxy for performer ratings during in-the-wild
  interaction.
\end{abstract}

@@

\begin{itemize}
\item we should be clear (and careful) where we're talking about
  groups TMs, and where we're talking about individual performers
\item what's the terminology we're using for a ``performance''? Is it
  ``performance'', or is that too music-specific? ``Session''?
  Alternatives are ``activity'', or ``interaction'', but they're so
  general\ldots This decision has implications for our other
  terminology: participant vs performer vs musician, etc.
\item Title: maybe ``Traces through time: making sense of longitudinal
  open-ended interaction data'' or some such? I think that's our big
  contribution: it takes the in-the-wild, practitioner-driven research
  program stuff and tries to make some quantitative sense out of it
  post-hoc.
\end{itemize}

@@

Ideas: the in-the-wild nature, and the different contexts - both the
fact that they are different, and the fact that their differences make
collecting certain data/responses easier or harder

@@

Suppose that each participant's activity in a collaborative
interaction can at any given time be classified as belonging to one of
$m$ distinct \emph{states}. Over the course of the activity, each
participant $X$ generates a length $N$ sequence of these interaction
states
\begin{equation}
 X_n \hskip 2em n = 1, \ldots, N
\end{equation}
The transitions between these interaction states can be summarised in
a transition matrix (TM) as if the sequence $X_n$ were a first-order Markov
chain. Let $N_{ij}$ be the number of times that state $j$ follows
state $i$. Each participant's TM $P$, will then be an
$m \times m$ matrix with entries
\begin{equation}
  p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}
\end{equation}
Each entry of $P$ will be such that the transition probability from
state $i$ to state $j$ is given by the entry in the $i^{th}$ row and
the $j^{th}$ column. The transition activity of the whole ensemble can
be calculated by averaging the TM for each performer.
The matrix can be normalised so that the sum of all elements is equal
to one.

If there is an (automated) classification process for determining each
participant's state sequence during an activity, then it is a simple
process to generate a TM for each participant and for the group as a
whole. The challenge is to understand what the TM reveals about the
group interaction, and to allow for comparisons between sessions and
groups.

To compare the TMs of different interactions we use two high-level
quantities, ``flux'' and ``entropy''. The flux of a sequence is
defined as the ratio of change transitions (e.g. $A \rightarrow B$
where $A \neq B$), to self transitions (e.g. $A \rightarrow A $). The
flux of a TM $P$ is given by
\begin{equation}
  \mathrm{flux}(P) = 1 - \mathrm{tr}(P)
\end{equation}
where the trace $\mathrm{tr}(P)$ is the sum of the diagonal entries of
$P$. Intuitively, flux is a measure of how frequently the participant/group
changes state. Flux returns a value in the interval $[0,1]$ and will
return 0 when participants never change state, and 1 when participants
never stay in the same state for two measurements in a row.

Another useful measure that can be used on a transition matrix is its
entropy, defined in the information-theoretic~\cite{Shannon:1948rt}
sense:
\begin{equation}
  H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})
\end{equation}
This entropy measure is small when the matrix is sparse, and largest
when each matrix element is equal. It offers a different perspective
on collaborative interactions than the flux measure by capturing the
breadth of the gestural space explored by the participants throughout
the course of an interaction. Consider the degenerate case of a
participant who alternates between two states for a whole performance:
the flux in this case will be maximal ($\mathrm{flux}(P) = 1$) since
there are no self-transitions, (only $A \rightarrow B$ or
$ B \rightarrow A$) even though the participant has only used a small
subset of the state space. The entropy of this interaction, on the
other hand, will be low. Entropy, therefore, is a measure of how broad
the participant's exploration of the state space is in a given
interaction.

% \section{Making sense of two year's worth of touch-screen
%   interactions}
\section{A corpus of touch-screen interaction}

@@

% TODO maybe a ref to the CHI paper here?

% TODO I think this ``scene-setting'' stuff could actually go in the
% first section of the paper, with the TM stuff following.  But I'm
% open to push-back on that.

Over a period of more than two years (April 2013 - June 2015), we have
collected data from $150$ group music-making sessions where
participants used musical touch-screen interfaces on Apple iPad
devices. These interfaces all shared a common mode of interaction
where the majority of the screen could be used for free form touch
interactions; tapping produced short sounds, while swiping or swirling
produced long sounds with volume proportional to the velocity of the
moving touch point. Through all of these performances every touch
interaction---every touch down, dragged and up---was timestamped and
logged to disk for later analysis.

These sessions were part of a long-term research project into music
touch-screen interaction, as well as an ongoing creative practice. As
a result, the sessions had a variety of contexts including rehearsals,
performances, lab study sessions, classes, and casual demonstrations,
some of which are shown in Figure
\ref{fig:three-performance-contexts}. Participants performed free
improvisations as well as written compositions and performed on iPads
alone or using setups of iPads and acoustic percussion instruments.
The median length of these sessions was 7m29s and the median number of
participants was four, other descriptive statistics about the sessions
is shown in Table \ref{corpus-table}. Each session was coded with
metadata regarding the session context (whether it was a performance,
rehearsal, study, demonstration, or data collection session), the
session type (either improvisation or composition), and the
instruments used (iPads alone, or iPads with acoustic percussion). The
distribution of recorded sessions by context and type is shown in
Figure \ref{fig:count-data}.

% TODO again, I think this bit would be beefed-up if we included the
% fact that it's already been published.  And it also means we don't
% have to go into as much detail about how we did it - the main thing
% is that we've got a sensible (and peer-reviewed!) way of taking the
% touch interaction data and generating gesture sequences (and
% therefore TMs)

To analyse the gestural interactions present in each of these
sessions, the sessions were automatically coded according to a
vocabulary of nine simple, continuous, touch-screen gestures. Examples
of these gestures was generated by an expert user in formal process
where each gesture was performed for 1-minute and feature vectors of
discriptive statistics on these examples was used to train a Random
Forest~\cite{Breiman:2001kx} classifier\footnote{The classifier used
  was from Python's Scikit-learn package~\cite{scikit-learn}.}. The
classifier was applied to 5-second windows of the touch-data each
second provide a chain of gesture-states for each participant in each
session. These chains can be used to create $9 \times 9$
gesture-transition matrices for individual performers, and the whole
ensemble in each session.

While the interaction sessions recorded in our corpus were performed
by many different participants with varying experience levels, in
different contexts and on different instrumental setups, the common
interface and method of recording the data allows all the sessions to
be transformed into gesture sequences. This common format allows us to
ask if performance styles vary in between these different situations.
The sequences can be plotted and inspected qualitatively, but
transition matrices present a quantitative approach to analysing such
interactions.

In the rest of this paper we will use our matrix measure techniques to
interrogate this dataset and to answer two questions. First, do the
transition matrix measures allow us to distinguish between different
performance contexts, types and instrumental setups? Secondly, do
these performance measure allow us further insight into the corpus
than the metadata can provide?

@@

The distribution of flux and entropy with respect to session contexts,
types and instrumentation are shown in Figure
\ref{fig:flux-entropy-boxplot}. A three-way ANOVA procedure was
performed on the data set to find whether the metatdata could
significantly predict the entropy and flux values.
% TODO we haven't explained what ``the 95 performative sessions'' are,
% have we?
In the following analysis, only the 95 performative sessions were
considered, excluding the demonstration and data-collection contexts.

@@

\begin{itemize}
\item TODO: need to add overall contrasts column to do pairwise t-tests.
\end{itemize}

@@

It is clear that flux and entropy can detect the variation in
performance styles in between performances of different types, but can
they be used to understand some of the internal structure of
performances? A simple model for understanding temporal art-forms is
the ``beginning, middle, end'' structure. To investigate how this
structure fits our corpus, we can divide the gestural-sequences of
each session into thirds and calculate the transition matrices given
by each section.

@@

% Determining the strengths of these positive relationships is
% difficult, both due to the problem of treating ordinal Likert
% responses as numeric~\cite{Gardner:2007fj} and also
% due to the unknown distribution of the entropy and flux statistics. As
% a result, we use a conservative proportional odds logistic regression
% (POLR)~(using the \texttt{polr} function in \texttt{R}, from the MASS
% package \cite{Venables:2002qv}) to treat the Likert responses as an
% ordered factor response.

>>>>

\title{Quantitative Evaluation of Touch-Screen Group Musical Performance with
  Transition Matrix Measures}
\title{Mining interaction log data in long-term practice-led research}

@@

\begin{abstract}
  Interfaces for supporting collaborative creativity are often
  evaluated with user ratings and interviews. Such data can be
  impractical to collect in-the-wild, for example during musical
  performances. In this paper we show how a transition matrix based
  post-hoc analysis of logged interaction data from a long program of
  experiments in group musical creativity can shed light on the nature
  of the group interaction in-the-wild. Using a corpus of more than
  140 performances, rehearsals, and demonstrations over more than two
  years, we find that two matrix measures, flux and entropy, can be
  used as a proxy for performer ratings during in-the-wild
  interaction.
\end{abstract}

@@

% \begin{itemize}
% \item we should be clear (and careful) where we're talking about
%   groups TMs, and where we're talking about individual performers
% \item what's the terminology we're using for a ``performance''? Is it
%   ``performance'', or is that too music-specific? ``Session''?
%   Alternatives are ``activity'', or ``interaction'', but they're so
%   general\ldots This decision has implications for our other
%   terminology: participant vs performer vs musician, etc.
% \item Title: maybe ``Traces through time: making sense of longitudinal
%   open-ended interaction data'' or some such? I think that's our big
%   contribution: it takes the in-the-wild, practitioner-driven research
%   program stuff and tries to make some quantitative sense out of it
%   post-hoc.
% \end{itemize}

@@

Computing devices and interfaces continue to open up rich, open-ended
interaction contexts such as group music-making. Understanding and
evaluating interfaces in these contexts presents many challenges;
making use of the easily-logged interaction data without descending
into ``quantitative reductionism'', as well as balancing the desire to
gather ``felt experience'' data through surveys/interviews without
unnecessary interruptions to the participants.

In this paper we show how we have used post-hoc analyses of log data
to support a long-term practice-led research program into iPad
interfaces for group music-making. Over a period of more than two
years (April 2013--June 2015), we have collected data from $150$ group
music-making sessions where participants used musical touch-screen
interfaces on Apple iPad devices. These interfaces (see\cite{TODO} for
details) shared a common mode of interaction where the majority of
the screen could be used for free form touch interactions; tapping
produced short sounds, while swiping or swirling produced long sounds
with volume proportional to the velocity of the moving touch point.
Through all of these performances every touch interaction---every
touch down, dragged and up---was timestamped and logged to disk for
later analysis.

These sessions were part of a long-term research project into music
touch-screen interaction, as well as an ongoing creative practice. The
log data collected over these sessions is homogeneous in form, but
represents a wide variety of performance contexts: rehearsals and
performances, written compositions and free improvisations, controlled
lab studies and casual demonstrations, with and without additional
acoustic instruments (some of these are shown in
Figure~\ref{fig:three-performance-contexts}).

The median length of these sessions was 7m29s and the median number of
participants was four; other descriptive statistics about the sessions
are shown in Table \ref{corpus-table}. The session data includes
metadata regarding the session \emph{context} (whether it was a
performance, rehearsal, study, demonstration, or data collection
session), session \emph{type} (either improvisation or composition),
and the instruments used (iPads alone, or iPads with acoustic
percussion). The distribution of recorded sessions by context and type
is shown in Figure \ref{fig:count-data}, and we shall use these
groupings for our analysis in
Section~\ref{differentiating-interaction-sessions}.

In some of these sessions, surveys and open-ended interviews were
conducted with the participating musicians in an effort to understand
the interaction dynamics and felt experience of the participants.
However, many of these sessions were in-the-wild musical performances,
and in these cases no such subjective evaluation data was
collected---only the log data. This raises the question: can we use
the log data to fill in some of the missing ``subjective experience''
data, and to understand broader trends in the interaction behaviour of
the musicians over this period?

@@

Suppose that each participant's activity in a collaborative
interaction can at any given time be classified as belonging to one of
$m$ distinct \emph{states}. Over the course of the activity, each
participant $X$ generates a length $N$ sequence of these interaction
states
\begin{equation}
 X_n \hskip 2em n = 1, \ldots, N
\end{equation}
which can be thought of as Markov process. If there is an
(automated) classification process for extracting each participant's
state sequence from (or during) their interaction data, then we can
estimate the $m \times m$ transition matrix $P$ which characterises
this process by simply counting the number of times state $j$ follows
state $i$
\begin{equation}
  p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}
\end{equation}
The transition activity of the whole ensemble can be calculated by
averaging the TM for each performer. The matrix can be normalised so
that the sum of all elements is equal to one.

From previous work\cite{TODO} we have a robust technique for
segmenting the raw touch data generated by the instrument into a
sequence of high-level ``gestures'' using a Random
Forest\cite{Breiman:2001kx} classifier\footnote{The classifier used
  was from Python's Scikit-learn package\cite{scikit-learn}.}. The
classifier was applied to 5-second windows of the touch-data to
provide a chain of gesture-states for each participant in each
session. These chains can be used to create $9 \times 9$ (the
classifier recognises 9~distinct gestures) transition matrices for
individual performers and the whole ensemble in each session.

\section{Quantitative transition matrix statistics}
\label{sec:underst-impr-group}

While the interaction sessions recorded in our corpus were performed
by many different participants with varying experience levels, in
different contexts and on different instrumental setups, the common
interface and method of recording the data allows all the sessions to
be transformed into gesture sequences. These transition matrices have
been used as a visualisation technique to reveal qualitative
differences in structure in improvisational
music-making\cite{Swift:2014tya}. In this paper, however, we wish to
extract quantitative statistics from these TMs for comparing and
tracking different sessions over time.

To do this we calculate two high-level quantities from each session:
``flux'' and ``entropy''. The flux of a sequence is defined as the
ratio of state transitions (e.g. $A \rightarrow B$ where $A \neq B$),
to self transitions (e.g. $A \rightarrow A $). The flux of a TM $P$ is
given by $\mathrm{flux}(P) = 1 - \mathrm{tr}(P)$ where the trace
$\mathrm{tr}(P)$ is the sum of the diagonal entries of $P$.
Intuitively, flux is a measure of how frequently the participant/group
changes state. Flux returns a value in the interval $[0,1]$ and will
return 0 when participants never change state, and 1 when participants
never stay in the same state for two measurements in a row.

Another useful measure that can be used on a transition matrix is its
entropy, defined in the information-theoretic\cite{Shannon:1948rt}
sense:
\begin{equation}
  H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})
\end{equation}
This entropy measure is small when the matrix is sparse, and largest
when each matrix element is equal. It offers a different perspective
on collaborative interactions than the flux measure by capturing the
breadth of the gestural space explored by the participants throughout
the course of an interaction. Consider the degenerate case of a
participant who alternates between two states for a whole performance:
the flux in this case will be maximal ($\mathrm{flux}(P) = 1$) since
there are no self-transitions, (only $A \rightarrow B$ or
$ B \rightarrow A$) even though the participant has only used a small
subset of the state space. The entropy of this interaction, on the
other hand, will be low. Entropy, therefore, is a measure of how broad
the participant's exploration of the state space is in a given
interaction.

In the rest of this paper we will use these matrix measures to answer
two questions. First, do the transition matrix measures allow us to
distinguish between different performance contexts, types and
instrumental setups? Secondly, do these performance measure allow us
further insight into the corpus than the metadata can provide? TODO I
think this second question is too broad?

@@

@@

The distribution of flux and entropy with respect to session contexts,
types and instrumentation are shown in Figure
\ref{fig:flux-entropy-boxplot}. A three-way ANOVA procedure was
performed on the data set to find whether the metatdata could
significantly predict the entropy and flux values.
% TODO we haven't explained what ``the 95 performative sessions'' are,
% have we?
In the following analysis, only the 95~``performance context''
sessions were considered, excluding the demonstration and
data-collection contexts. TODO can we justify this?

@@

@@

It is clear from these results that flux and entropy can be used to
differentiate between performance styles and types, but can they be
used to understand some of the internal structure of performances? A
simple model for understanding temporal art-forms is the ``beginning,
middle, end'' structure. To investigate how this structure fits our
corpus, we can divide the gestural-sequences of each session into
thirds and calculate the transition matrices given by each section.

@@

% Determining the strengths of these positive relationships is
% difficult, both due to the problem of treating ordinal Likert
% responses as numeric\cite{Gardner:2007fj} and also
% due to the unknown distribution of the entropy and flux statistics. As
% a result, we use a conservative proportional odds logistic regression
% (POLR)~(using the \texttt{polr} function in \texttt{R}, from the MASS
% package \cite{Venables:2002qv}) to treat the Likert responses as an
% ordered factor response.
